---
title: "Practical Machine Learning - Course Project"
author: "Travis Hill"
date: "July 19, 2016"
output: html_document
---

# Introduction

#### Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

#### Data

The training data for this project are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

#### Goal

The goal of your project is to predict the manner in which they did the exercise. This is the “classe” variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

# Preload Necessary Packages

```{r warning=FALSE, echo=TRUE, error=FALSE, message=FALSE}
library(caret)
library(knitr)
library(randomForest)
library(rattle)
library(RColorBrewer)
library(rpart)
library(rpart.plot)
```

# Loading and Cleaning Data

#### Obtain and Load Data

```{r, echo=TRUE}
set.seed(12345)
testingUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainingUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testing <- read.csv(url(testingUrl), na.strings=c("NA","#DIV/0!",""))
training <- read.csv(url(trainingUrl), na.strings=c("NA","#DIV/0!",""))

# Partition the training set into two sets.

inTrain <- createDataPartition(training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]
myTesting <- training[-inTrain, ]
```

Check the dimensions of the two new sets sets.

```{r, echo=TRUE}
dim(myTraining)
dim(myTesting)
```

#### Cleaning the Data

First, remove `NearZeroVariance` variables, and remove first column of `myTraining` data.

```{r, echo=TRUE}
nzv <- nearZeroVar(myTraining, saveMetrics=TRUE)
myTraining <- myTraining[,nzv$nzv==FALSE]
nzv<- nearZeroVar(myTesting,saveMetrics=TRUE)
myTesting <- myTesting[,nzv$nzv==FALSE]

myTraining <- myTraining[c(-1)]
```

Next, remove variables with over 60% `NA` values.

```{r, echo=TRUE}
trainingTemp <- myTraining
for(i in 1:length(myTraining)) {
    if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .7) {
        for(j in 1:length(trainingTemp)) {
            if( length( grep(names(myTraining[i]), names(trainingTemp)[j]) ) == 1)  {
                trainingTemp <- trainingTemp[ , -j]
            }   
        } 
    }
}
myTraining <- trainingTemp
rm(trainingTemp)
```

Finally, we transfer both the `myTesting` and `testing` datasets, allowing only variables in testing sets that are featured in training set, and coerce the data to the same type between `testing` and `myTraining` sets.

```{r, echo=TRUE}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58])  # removing the classe column
myTesting <- myTesting[clean1]        
testing <- testing[clean2]       

#Coerce the data sets to same type

for (i in 1:length(testing) ) {
    for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testing)[j]) ) == 1)  {
            class(testing[j]) <- class(myTraining[i])
        }      
    }      
}

testing <- rbind(myTraining[2, -58] , testing)
testing <- testing[-1,]

#Check dimensions of data sets one more time

dim(myTesting)
dim(testing)
```

# Prediction Using Decision Trees
```{r, echo=TRUE}
set.seed(12345)
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
fancyRpartPlot(modFitA1)

predictionsA1 <- predict(modFitA1, myTesting, type = "class")
cmtree <- confusionMatrix(predictionsA1, myTesting$classe)
cmtree

plot(cmtree$table, col = cmtree$byClass, main = paste("Decision Tree Confusion Matrix: Accuracy =", round(cmtree$overall['Accuracy'], 4)))
```

# Prediction Using Random Forests
```{r, echo=TRUE}
set.seed(12345)
modFitB1 <- randomForest(classe ~ ., data=myTraining)
predictionB1 <- predict(modFitB1, myTesting, type = "class")
cmrf <- confusionMatrix(predictionB1, myTesting$classe)
cmrf

plot(modFitB1)

plot(cmrf$table, col = cmtree$byClass, main = paste("Random Forest Confusion Matrix: Accuracy =", round(cmrf$overall['Accuracy'], 4)))
```

# Prediction on Test Data (for Submission)

After running both models, Random Forests gave a higher accuracy than Decision Trees, with an accuracy from the `myTesting` dataset of `99.89%`, resulting in an expected out of sample error of `0.11%`. Given this, we will use Random Forests to predict against the `testing` set.

```{r, echo=TRUE}
predictionB2 <- predict(modFitB1, testing, type = "class")
predictionB2
```

Finally, we'll write these to text to submit to Coursera.

```{r, echo=TRUE}
writeFiles = function(x){
    n = length(x)
    for(i in 1:n){
        filename = paste0("problem_id_",i,".txt")
        write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
}

#writeFiles(predictionB2)
# execution commented out to protect graders, submission text files were initially created.
```

